Exploring probability distributions for bivariate temporal granularities

### 2019-12-10 main.Rmd
language change in cricket analysis

The application is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized in other applications. The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. With eight teams, each team plays each other twice in a home-and-away round-robin format in the league phase. In a Twenty20 game the two teams have a single innings each, which is restricted to a maximum of 20 overs. Hence, in this format of cricket, a match will consist of 2 innings, an innings will consist of 20 overs, an over will consist of 6 balls with some exceptions.

The ball by ball data for IPL season 2008 to 2016 is fetched from [Kaggle](https://www.kaggle.com/josephgpinto/ipl-data-analysis/data).  The `cricket` data set in the `gravitas` package summarizes the ball-by-ball data cross overs and contains information for a sample of 214 matches spanning 9 seasons (2008 to 2016).

@@

Although there is no conventional time granularity in cricket, we can still represent the data set `cricket` through a `tsibble`, where each over, which represents an ordering from past to future, can form the index of the tsibble. The hierarchy table would look like the following:

```{r hierarchy2}
library(gravitas)
library(tibble)
hierarchy_model <- tibble::tibble(
  G = c("over", "inning", "match"),
  k = c(20, 2, 1))

knitr::kable(hierarchy_model, format = "markdown")

@@

First, we look at the distribution of runs per over across over of the innings and seasons in \autoref{fig:seas-over-inning}. The distribution of runs per over has not significantly changed from 2008 to 2016. There is no clear pattern/trend that runs per over is increasing or decreasing across seasons. Hence, we work with subsets of seasons to answer some of the questions:

@@

Mumbai Indians(MI) and Chennai Super kings(CSK) are considered one of the best teams in IPL with multiple winning titles and always appearing in final 4 from 2010 to 2015. It would be interesting to take their example in order to dive deeper into the first question.

@@

Q2: Is run rate set to reduce in subsequent over if fielding/bowling is good in the previous over? Between fielding and bowling which penalizes runs in the subsequent over more?

For establishing that the fielder fielded well in a particular over, we can see how many catches and run outs were made in that particular over. If a batsman is bowled out, it does not necessarily signify good fielding. So we only include catches and run out as a measure of fielding. Difference in runs across over should be negative if good fielding has an impact on the runs scored in the subsequent overs. Let us see if this fact is true. Figure \ref{fig:field} shows the difference between run rate between two subsequent overs are negative when good fielding leads to one dismissal in an over, implying good fielding in one over has indeed an impact on runs scored in the subsequent overs.

@@

Q3: Is runs set to reduce in the next over for dot balls in this over?

@@

>>>>

The application is not only restricted to temporal data. We provide an example of cricket to illustrate how this can be generalized in other applications. The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. In a Twenty20 game, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season can consist of several matches. The ball by ball data for IPL season 2008 to 2016 is fetched from [Kaggle](https://www.kaggle.com/josephgpinto/ipl-data-analysis/data). The `cricket` data set in the `gravitas` package summarizes the ball-by-ball data across overs and contains information for a sample of 214 matches spanning 9 seasons (2008 to 2016), each of which has 6 balls in an over and 2 innings in a match.
s

<!-- the two teams have a single innings each, which is restricted to a maximum of 20 overs. Hence, in this format of cricket, a match will consist of 2 innings, an innings will consist of 20 overs, an over will consist of 6 balls with some exceptions. -->

@@

Although there is no conventional time granularity in cricket, we can still represent the data set `cricket` through a `tsibble`, with each over representing an ordering from past to future, can form the index of the tsibble. The hierarchy table would look like the following:

```{r hierarchy2}
library(gravitas)
library(tibble)
hierarchy_model <- tibble::tibble(
  G = c("over", "inning", "match", "season"),
  C = c("over-of-inning", "inning-of-match", "match-of-season",1),
  P = c(20, 2, "aperiodic", 1))
knitr::kable(hierarchy_model, format = "markdown")

@@

First, we look at the distribution of runs (measured variable) across over-of-inning (circular granularity) and match-of-season (aperidic cyclic granularity) in \autoref{fig:seas-over-inning}. The distribution of runs per over has not significantly changed from 2008 to 2016. There is no clear pattern/trend that runs per over is increasing or decreasing across seasons. Hence, we work with subsets of seasons to answer some of the questions:

@@

Mumbai Indians (MI) and Chennai Super kings (CSK) are considered one of the best teams in IPL with multiple winning titles and always appearing in final playoffs from 2010 to 2015. It would be interesting to take their example in order to dive deeper into the first question.

@@

Q2: Is runs per over set to reduce in subsequent over if fielding is good in the previous over?

For establishing that the fielder fielded well in a particular over, we can see how many catches and run outs were made in that particular over. If a batsman is bowled out, it does not necessarily signify good fielding. So we only include number of catches and run out in an over as a measure of good fielding. Difference in runs across over should be negative if good fielding has an impact on the runs scored in the subsequent overs. We can verify this fact through Figure \ref{fig:field} which shows that the difference in runs between two subsequent overs are negative when good fielding leads to one dismissal in an over, implying good fielding in one over has indeed an impact on runs scored in the subsequent overs.

@@

Q3: Are runs set to reduce in the next over for dot balls in the previous over?

@@

In Q2 and Q3, we considered dots balls per over or catches per over as aperiodic cyclic granularities. They do not appear in the hierarchy table since we can not compute these granularities as a function of circular or quasi-circular granularities. Also, their order in the hierarchy table is not clear as they are essentially obtained from some aperiodic linear granularity.
